Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra

### 2018-11-02 manuscript/manuscript.tex
Polish the manuscript.

@@

\title{Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra}
\author[1]{Daniel P. Rice}
\author[1,*]{John Novembre}
\author[2,*]{Michael M. Desai}
\affil[1]{Department of Human Genetics, University of Chicago, Chicago, IL}
\affil[2]{Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA}
\affil[*]{These authors contributed equally to this work.}
\date{\today}

@@

\abstract{
Demographic inference methods in population genetics typically assume that the ancestry of a sample can be modeled by the Kingman coalescent.
A defining feature of this stochastic process is that it generates genealogies that are binary trees: no more than two ancestral lineages may coalesce at the same time.
However, this assumption breaks down under several scenarios.
For example, pervasive natural selection and extreme variation in offspring number can both generate genealogies with ``multiple-merger'' events in which more than two lineages coalesce instantaneously.
Therefore, detecting multiple mergers is important both for understanding which forces have shaped the diversity of a population and for avoiding fitting misspecified models to data.
Current methods to detect multiple mergers in genomic data rely on the site frequency spectrum (SFS).
However, the signatures of multiple mergers in the SFS are also consistent with a Kingman coalescent with a time-varying population size.
Here, we present a new method for detecting multiple mergers based on the pointwise mutual information of the two-site frequency spectrum for pairs of linked sites.
Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than on their branch lengths and is therefore largely insensitive to population size change.
This statistic is global in the sense that it can detect when the genome-wide genetic diversity is inconsistent with the Kingman coalescent, rather than detecting outlier regions, as in selection scan methods.}

@@

While the example above is extreme, it is well-established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
For example, \cite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
In a similar vein, \cite{CvijovicEtAl2018}, showed that reduction of genetic diversity by purifying selection is accompanied by distortions in the SFS, leading to a false signal of population growth.
Moreover, genomic evidence from multiple species suggests that such violations of neutrality may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

@@

One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
To this end, \cite{BirknerEtAl2013, BlathEtAl2016} and \cite{SpenceEtAl2016} derived methods for computing the expected SFS of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the SFS is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.

However, existing methods are limited in their ability to distinguish multiple mergers from general models of population-size change.
The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
\cite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but identifying that the SFS increases at high frequencies requires knowing the ancestral allele at each site.
High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS.

Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
These statistics may be calculated efficiently from genomic single nucleotide variant data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.

Following the notation of \cite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complementary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without knowing the physical locations of the SNPs.

In contrast, the 2-SFS is a statistic of \textit{pairs} of sites.
We define the 2-SFS,
$\left\{ \xi_{ij}(d) : d > 0; 1 \leq i, j < n\right\}$,
as the fraction of pairs of sites separated by $d$ bases for which there is a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
The 2-SFS has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and by \cite{Xie2011a} in a model with selection.
We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies. (For non-recombining sites, the 2-SFS is independent of the distance and so we will suppress the $d$ in our notation.)

@@

\cite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite sites locus under the standard time-homogeneous Kingman coalescent.
He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \not\in \{i, (n-i)\}$.
This result, combined with \eqs{eq:expected_sfs}{eq:expected_2sfs}, implies a negative correlation between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \not\in \{i, (n-i)\}$.
(There are positive correlations between mutations at complementary frequencies induced by genealogies whose root node partitions the tree into subtrees of size $i$ and $n-i$.)

\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, ranges from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in populations that are rapidly adapting or experiencing extensive purifying selection \autocite{NeherHallatschek2013}.
The calculations of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6).
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

In the following, we demonstrate that the positive associations between mutations at different frequencies distinguish the multiple-merger from the Kingman coalescent and that this distinction
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
    \item applies to Kingman coalescents with time-varying coalescent rates
    \item is robust to recombination between the sites
    \item is also a feature of forward-time models with selection
    \item can form the basis of a model-checking procedure for demographic inference methods
\end{enumerate*}.
To demonstrate these four properties, we introduce a transformation of the 2-SFS that we call frequency pointwise mutual information ($\fpmi$).
We use a combination of numerical calculations and stochastic simulations to explore the properties of this statistic under different coalescent models.
Finally, we demonstrate the use of $\fpmi$ in a graphical model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

@@

We aim to quantify the dependence between minor allele counts at a pair of sites, particularly for $i\neq j$.
To do so, we define the \textit{frequency pointwise mutual information} as
\begin{align}
    \fpmi(i,j;d) = \log \frac{\E{\eta_{i,j}(d)}}{\E{\eta_i} \E{\eta_j}}.
    \label{eq:fpmi}
\end{align}
\Fig{fig:schematic} shows the steps to compute $\fpmi$ from a sample of sequences.

\begin{figure}
\centering
\includegraphics[scale=1]{figures/schematic.pdf}
\caption{Summarizing population genetic data with frequency pointwise information ($\fpmi$). First, a sequence alignment is reduced to the minor allele frequencies and positions of segregating mutations. Then, for a given distance, $d$, the 2-SFS gives the number of pairs of sites with each combination of minor allele frequencies. Finally, $\fpmi$ is calculated from the 2-SFS according to \eq{eq:fpmi}.
 \label{fig:schematic}}
\end{figure}

@@

This transformation of the 2-SFS has several useful properties.
First, because $\fpmi$ is based on the minor allele frequencies, we may compute it without knowing the ancestral allele.
Second, the results of \cite{Fu1995} for the folded SFS imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
Finally, $\fpmi$ normalizes for distortions in the coalescence time distribution reflected in the single site SFS.
In the Kingman coalescent, population-size variation alters the coalescence time distribution but preserves the distribution of tree topologies.
We will show that, because $\fpmi$ normalizes the 2-SFS by the product of single site spectra, it is largely insensitive to the distribution of coalescence times.
Thus, $\fpmi$ primarily reflects the distribution of topologies, which distinguishes multiple mergers from the Kingman coalescent.

@@

With finite data, estimates of the 2-SFS will be noisy.
This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman \autocite{Fu1995} and faster with growth or multiple mergers.
We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
Thus, to detect multiple mergers, we use a binned form of the SFS and 2-SFS:
\begin{align}
    \eta_{\text{lo}}(i_c) &= \sum_{i=1}^{i_c-1} \eta_i \\
    \eta_{\text{hi}}(i_c) &= \sum_{i=i_c}^{\floor{n/2}} \eta_i \\
    \eta_{\text{hi,lo}}(d; i_c) &= \sum_{i=1}^{i_c-1} \sum_{j=i_c}^{\floor{n/2}} \eta_{ij}(d),
\end{align}
where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
Binning provides a stable estimate of the 2-SFS for large sample sizes because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
We note that a less coarse-grained binning scheme may strike a balance between sampling noise and preserving more detailed information about allele frequencies.

@@

We implemented numerical computations of the moments of the branch lengths $\E{\tau_i}$ and $\E{\tau_{ij}}$.
For the Kingman coalescent with time-varying coalescent rate, we used equations~(1)-(12) of \cite{ZivkovicWiehe2008}.
For the beta coalescent, we implemented the recursion described by \cite{BirknerEtAl2013}.

Functions for computing the branch-length moments are implimented in \texttt{python} using the numerical package \texttt{numpy} and available [GITHUB].
Our Kingman coalescent code can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability described in \cite{GriffithsTavare1994}.
Thus, these formulas are only practical for sample sizes up to $n \approx 40$.
The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ and is also only practical for samples up to $n \approx 50$.

@@

We ran coalescent simulations using a custom version of \texttt{msprime} \autocite{KelleherEtAl2016} capable of multiple mergers, based on modifications made by Joe Zhu [GITHUB].
This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available [GITHUB].

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d\,r$.
(See SI for parameter combinations.)
We chose values of $d\,r$ so that the realized values of $d\,r \E{T_2}$ varied over several orders of magnitude.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at the ends of the locus.
We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{Messer2013}.
In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and per basepair mutation rate $\mu = 10^{-7}$.
We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of simulating a finite chromosome.

In order to vary the effects of sweeps on neutral diversity, we varied the fraction of mutations that are beneficial $f_{\text{sel}}$ over several orders of magnitude: $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$.
For each $f_{\text{sel}}$, we ran 100 independent replicate simulations and computed $\eta_i$ and $\eta_{ij}$ of neutral mutations averaged over all replicates.

\texttt{SLiM} parameter files, \texttt{python} wrapper scripts for parsing output, and \texttt{snakemake} files for running simulations are available [GITHUB].

@@

We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtAl2011}.
The SNP calls that characterize these sequences were subjected to a variety of quality filters described in \cite{LackEtAl2015}.
We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
These files contain the sequence alignments of all flies in the sample on all chromosome arms.
We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by selecting the first 100 remaining samples in alphanumeric order by sample name.
As a result, the data for each chromosome arm is from a slightly different subset of the individuals.

We calculated the average pairwise diversity, $\Pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres.
Our modeling---and coalescent-based demographic inference in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous values of $\Pi$ for further analysis.
The boundary positions of these central regions are given in Table~\ref{tab:central_regions}.

In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites with fewer than 90 of the 100 genotypes called.
This leaves over 90\% of all sites and without substantially altering the fraction of polymorphic sites (Table~\ref{tab:called_sites}).

We fit a demographic model to the folded SFS of fourfold degenerate sites for each chromosome arm separately using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
Following \cite{RagsdaleGutenkunst2017} we fit a three-epoch piecewise constant--$N$ model:
\begin{align}
    N(t) = \begin{cases}
                N_1 & 0 \leq t \leq t_1 \\
                N_2 & t_1 <  t \leq t_2 \\
                N_{\text{anc}} & t > t_2,
            \end{cases}
    \label{eq:piecewise}
\end{align}
estimating both change-points ($t_1 < t_2$) and population sizes ($N_1$, $N_2$).
We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \cite{RagsdaleGutenkunst2017}.
For all four chromosomes, \texttt{fastNeutrino} inferred similar population growth.
Fitted parameters are presented in Table~\ref{tab:DPGP_params}.
We simulated the SFS and 2-SFS under the fitted parameters using \texttt{msprime}.

>>>>

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

@@

\title{Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra}
\author[1,2]{Daniel P. Rice}
\author[1,2,*]{John Novembre}
\author[3,4,*]{Michael M. Desai}
\affil[1]{Department of Human Genetics, University of Chicago, Chicago, IL}
\affil[2]{Department of Ecology and Evolution, University of Chicago, Chicago, IL}
\affil[3]{Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA}
\affil[4]{Department of Physics, Harvard University, Cambridge, MA}
\affil[*]{These authors contributed equally to this work.}
\date{\today}

@@

\abstract{
Demographic inference methods in population genetics typically assume that the ancestry of a sample can be modeled by the Kingman coalescent.
A defining feature of this stochastic process is that it generates genealogies that are binary trees: no more than two ancestral lineages may coalesce at the same time.
However, this assumption breaks down under several scenarios.
For example, pervasive natural selection and extreme variation in offspring number can both generate genealogies with ``multiple-merger'' events in which more than two lineages coalesce instantaneously.
Therefore, detecting multiple mergers is important both for understanding which forces have shaped the diversity of a population and for avoiding fitting misspecified models to data.
Current methods to detect multiple mergers in genomic data rely on the site frequency spectrum (SFS).
However, the signatures of multiple mergers in the SFS are also consistent with a Kingman coalescent with a time-varying population size.
Here, we present a new method for detecting multiple mergers based on the pointwise mutual information of the two-site frequency spectrum for pairs of linked sites.
Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than on their branch lengths and is therefore largely insensitive to population size change.
This statistic is global in the sense that it can detect when the genome-wide genetic diversity is inconsistent with the Kingman coalescent, rather than detecting outlier regions, as in selection scan methods.}
Finally, we demonstrate a graphical model-checking procedure based on the pointwise mutual information using genomic diversity data from \textit{Drosophila melanogaster}.

@@

While the example above is extreme, it is well established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
For example, \textcite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
In a similar vein, \textcite{CvijovicEtAl2018}, showed that reduction of genetic diversity by purifying selection is accompanied by distortions in the SFS, leading to a false signal of population growth.
Moreover, genomic evidence from multiple species suggests that such violations of neutrality may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

@@

One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
To this end, \textcite{BirknerEtAl2013, BlathEtAl2016} and \textcite{SpenceEtAl2016} derived methods for computing the expected SFS of (simultaneous-)multiple merger coalescents.
Further, \textcite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
In a related approach, \textcite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the SFS is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.

However, existing methods are limited in their ability to distinguish multiple mergers from general models of population-size change.
The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
\textcite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
The non-mononotic SFS identified by \textcite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but identifying that the SFS increases at high frequencies requires knowing the ancestral allele at each site.
High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS.

Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018}---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
These statistics may be calculated efficiently from genomic single-nucleotide-variant data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.

Following the notation of \textcite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complementary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without knowing the physical locations of the SNPs.

In contrast, the 2-SFS is a statistic of \textit{pairs} of sites.
We define the 2-SFS,
$\left\{ \xi_{ij}(d) : d > 0; 1 \leq i, j < n\right\}$,
as the fraction of pairs of sites separated by $d$ bases for which there is a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
The 2-SFS has been studied for a non-recombining sites by \textcite{FerrettiEtAl2018} in a neutral model and by \textcite{Xie2011a} in a model with selection.
We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies. (For non-recombining sites, the 2-SFS is independent of the distance and so we will suppress the $d$ in our notation.)

@@

\textcite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite-sites locus under the standard time-homogeneous Kingman coalescent.
He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \not\in \{i, (n-i)\}$.
This result, combined with \eqs{eq:expected_sfs}{eq:expected_2sfs}, implies a negative correlation between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \not\in \{i, (n-i)\}$.
(There are positive correlations between mutations at complementary frequencies induced by genealogies whose root node partitions the tree into subtrees of size $i$ and $n-i$.)

\textcite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, ranges from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in populations that are rapidly adapting or experiencing extensive purifying selection \autocite{NeherHallatschek2013}.
The calculations of \textcite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6 of \cite{BirknerEtal2013}).
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

In the following, we demonstrate that the positive associations between mutations at different frequencies distinguish the multiple-merger from the Kingman coalescent and that this distinction
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
    \item applies to Kingman coalescents with time-varying coalescent rates
    \item is robust to recombination between the sites
    \item is also a feature of forward-time models with selection
    \item can form the basis of a model-checking procedure for demographic inference methods
\end{enumerate*}.
To demonstrate these four properties, we introduce a transformation of the 2-SFS that we call frequency pointwise mutual information ($\fpmi$).
We use a combination of numerical calculations and stochastic simulations to explore the properties of this statistic under different coalescent models.
Finally, we demonstrate a graphical model-checking procedure based on $\fpmi$ using genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

@@

We aim to quantify the dependence between minor allele counts at a pair of sites, particularly for sites with different counts ($i\neq j$).
To do so, we define the \textit{frequency pointwise mutual information} as
\begin{align}
    \fpmi(i,j;d) = \log \frac{\E{\eta_{i,j}(d)}}{\E{\eta_i} \E{\eta_j}}.
    \label{eq:fpmi}
\end{align}
\Fig{fig:schematic} shows the steps to compute $\fpmi$ from a sample of sequences.

\begin{figure}
\centering
\includegraphics[scale=1]{figures/schematic.pdf}
\caption{Summarizing population genetic data with frequency pointwise mutual information ($\fpmi$). First, a sequence alignment is reduced to the minor allele frequencies and positions of segregating mutations. Then, for a given distance, $d$, the 2-SFS gives the number of pairs of sites with each combination of minor allele frequencies. Finally, $\fpmi$ is calculated from the 2-SFS according to \eq{eq:fpmi}.
 \label{fig:schematic}}
\end{figure}

@@

This transformation of the 2-SFS has several useful properties.
First, because $\fpmi$ is based on the minor allele frequencies, we may compute it without knowing the ancestral allele.
Second, the results of \textcite{Fu1995} for the folded SFS imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
Finally, $\fpmi$ normalizes for distortions in the coalescence time distribution reflected in the single site SFS.
In the Kingman coalescent, population-size variation alters the coalescence time distribution but preserves the distribution of tree topologies.
We will show that, because $\fpmi$ normalizes the 2-SFS by the product of single site spectra, it is largely insensitive to the distribution of coalescence times.
Thus, $\fpmi$ primarily reflects the distribution of topologies, which distinguishes multiple mergers from the Kingman coalescent.

@@

With finite data, estimates of the 2-SFS will be noisy.
This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman coalescent \autocite{Fu1995} and faster than this in models with population size growth or multiple mergers.
We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
Thus, to detect multiple mergers, we use a binned form of the SFS and 2-SFS:
\begin{align}
    \eta_{\text{lo}}(i_c) &= \sum_{i=1}^{i_c-1} \eta_i \\
    \eta_{\text{hi}}(i_c) &= \sum_{i=i_c}^{\floor{n/2}} \eta_i \\
    \eta_{\text{hi,lo}}(d; i_c) &= \sum_{i=1}^{i_c-1} \sum_{j=i_c}^{\floor{n/2}} \eta_{ij}(d),
\end{align}
where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
Binning provides a stable estimate of the 2-SFS for large sample sizes because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
We note that a less coarse-grained binning scheme may strike a balance between sampling noise and preserving more detailed information about allele frequencies.

@@

We implemented numerical computations of the moments of the branch lengths $\E{\tau_i}$ and $\E{\tau_{ij}}$.
For the Kingman coalescent with time-varying coalescent rate, we used equations~(1)-(12) of \textcite{ZivkovicWiehe2008}.
For the beta coalescent, we implemented the recursion described by \textcite{BirknerEtAl2013}.

Functions for computing the branch-length moments are implemented in \texttt{python} and available in the \texttt{git} repository for this project (\url{https://github.com/dp-rice/multiplemergers}).
Our Kingman coalescent code can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
The formulas of \textcite{ZivkovicWiehe2008} exhibit numerical instability related to the instability described in \textcite{GriffithsTavare1994}.
Thus, these formulas are only practical for sample sizes up to $n \approx 40$.
The recursion of \textcite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ and is also only practical for samples up to $n \approx 50$.

@@

We ran coalescent simulations using a custom version of \texttt{msprime} \autocite{KelleherEtAl2016} capable of multiple mergers, based on modifications made by Joe Zhu (\url{https://github.com/shajoezhu}).
This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available at \url{https://github.com/dp-rice/multiplemergers}.

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per-basepair recombination rate $r$ resulting in a per-locus total recombination rate $d\,r$.
(See SI for parameter combinations.)
We chose values of $d\,r$ so that the realized values of $d\,r \E{T_2}$ varied over several orders of magnitude.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at the ends of the locus.
We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{Messer2013}.
In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per-basepair $r = 10^{-8}$ and per-basepair mutation rate $\mu = 10^{-7}$.
We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of simulating a finite chromosome.

In order to vary the effects of sweeps on neutral diversity, we varied the fraction of mutations that are beneficial, $f_{\text{sel}}$, over several orders of magnitude: $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$.
For each $f_{\text{sel}}$, we ran 100 independent replicate simulations and computed $\eta_i$ and $\eta_{ij}$ of neutral mutations averaged over all replicates.

\texttt{SLiM} parameter files, \texttt{python} wrapper scripts for parsing output, and \texttt{snakemake} files for running simulations are available at \url{https://github.com/dp-rice/multiplemergers}.

@@

We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \textcite{LangleyEtAl2011}.
The SNP calls that characterize these sequences were subjected to a variety of quality filters described in \textcite{LackEtAl2015}.
We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
These files contain the sequence alignments of all flies in the sample on all chromosome arms.
We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by selecting the first 100 remaining samples in alphanumeric order by sample name.
As a result, the data for each chromosome arm is from a slightly different subset of the individuals.

We calculated the average pairwise diversity, $\Pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres, in agreement with calculations by \textcite{Corbett-DetigEtAl2015}.
Our modeling---and coalescent-based demographic inference in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous values of $\Pi$ for further analysis.
The boundary positions of these central regions are given in Table~\ref{tab:central_regions}.

In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites with fewer than 90 of the 100 genotypes called.
This leaves over 90\% of all sites and does not substantially alter the fraction of polymorphic sites (Table~\ref{tab:called_sites}).

We fit a demographic model to the folded SFS of fourfold degenerate sites for each chromosome arm separately using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
Following \textcite{RagsdaleGutenkunst2017} we fit a three-epoch piecewise constant--$N$ model:
\begin{align}
    N(t) = \begin{cases}
                N_1 & 0 \leq t \leq t_1 \\
                N_2 & t_1 <  t \leq t_2 \\
                N_{\text{anc}} & t > t_2,
            \end{cases}
    \label{eq:piecewise}
\end{align}
estimating both change-points ($t_1 < t_2$) and population sizes ($N_1$, $N_2$).
We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \textcite{RagsdaleGutenkunst2017}.
For all four chromosomes, \texttt{fastNeutrino} inferred similar population growth.
Fitted parameters are presented in Table~\ref{tab:DPGP_params}.
We simulated the SFS and 2-SFS under the fitted parameters using \texttt{msprime}.
